---
title: "Introduction to richStudio"
author:
  - name: Junguk Hur
    affiliation: University of North Dakota
    email: hurlabshared@gmail.com
  - name: Sarah Hong
    affiliation: Columbia University
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to richStudio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

**richStudio** is an interactive Shiny application for integrative functional
enrichment analysis and visualization. It provides researchers with tools to:
1. Perform functional enrichment analysis (GO, KEGG, Reactome)
2. Cluster functionally related terms using multiple algorithms
3. Visualize results through interactive plots
4. Save and reload analysis sessions

This vignette demonstrates the programmatic interface to richStudio's core
functions. For the interactive Shiny application, use `launch_richStudio()`.

# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("richStudio")
```
# Loading the Package

```{r load-package}
library(richStudio)
```

# Core Functions

## Merging Enrichment Results

The `merge_genesets()` function combines multiple enrichment results for
comparative analysis. Each input should be a data frame with columns including
`Term`, `GeneID`, `Pvalue`, and `Padj`.
```{r merge-example}
# Load example enrichment results
go1 <- read.delim(system.file("extdata", "go1.txt", package = "richStudio"))
go2 <- read.delim(system.file("extdata", "go2.txt", package = "richStudio"))
kegg1 <- read.delim(system.file("extdata", "kegg1.txt", package = "richStudio"))

# View structure of input data
head(go1[, c("Term", "Pvalue", "Padj")], 3)

# Merge multiple enrichment results
genesets <- list(GO_Sample1 = go1, GO_Sample2 = go2, KEGG_Sample = kegg1)
merged <- merge_genesets(genesets)

# View merged result structure
dim(merged)
head(merged[, c("Term", "Pvalue", "Padj")], 3)
```

## Clustering Enrichment Results

The `perform_clustering()` function groups functionally related terms. Three
methods are available:

1. **richR**: Kappa-based clustering (original method)
2. **hierarchical**: Hierarchical clustering with flexible linkage
3. **david**: DAVID-style functional clustering

```{r clustering-example}
# Perform clustering using the richR method
params <- list(cutoff = 0.5, overlap = 0.5, minSize = 2)
result <- perform_clustering(
  merged_gs = merged,
  method = "richR",
  params = params,
  gs_names = names(genesets)
)

# View clustering results
if (nrow(result$cluster_df) > 0) {
  cat("Number of clusters:", length(unique(result$cluster_df$Cluster)), "\n")
  cat("Total terms clustered:", nrow(result$cluster_df), "\n")
  head(result$cluster_summary, 5)
} else {
  cat("No clusters found with current parameters.\n")
  cat("Try adjusting the cutoff or overlap thresholds.\n")
}
```

## Available Clustering Methods

Use `get_clustering_methods()` to see which methods are available:

```{r methods}
methods <- get_clustering_methods()
for (name in names(methods)) {
  cat(sprintf("- %s: %s (Available: %s)\n",
              name,
              methods[[name]]$description,
              methods[[name]]$available))
}
```

# Launching the Interactive Application

For full functionality including visualization and session management,
launch the interactive Shiny application:

```{r launch, eval=FALSE}
launch_richStudio()
```

The application provides:

- **Home tab**: Welcome page with instructions
- **Enrichment tab**: Upload DEGs and run enrichment analysis
- **Clustering tab**: Select method, set parameters, and cluster terms
- **Visualization tab**: Interactive bar plots, dot plots, heatmaps, networks
- **Save/Load tab**: Session persistence and result export

# Session Information

```{r session-info}
sessionInfo()
```
